### 2019-09-20 submission/manuscript.Rmd
First full draft

## Abstract
75 words

@@

3000 words total

@@

***Failed experiments.*** If we never failed, we would not be trying hard enough. Over the past decade we have tried a number of experiments to improve the usability and utility of mothur. One of our first experiments was to use mothur to generate standard vector graphic (SVG)-formatted files of heatmaps and venn diagrams depicting the overlap between microbial communities. I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote. Such visuals require far too much customization to be publication-quality. Although QIIME has incorporated visualization tools through the Emperor package, the challenge of users taking default values has downsides as ordinations with black background or publishing 3-D ordinations in a 2-D medium litter the literature. Instead, we have encouraged users to use R packages to visualize mothur-generated results using the minimalR instructional materials. A second experiment we attempted was to create a graphical user interface (GUI) for running mothur. Forcing users to interact with mothur through the command line has been a significant hurdle for many. Unfortunately, the development effort required to create and maintain a GUI is significant and there is limited funding for such efforts. The newest version of QIIME (version 2) has emphasized interaction with the tools through a GUI and it remains to be seen how this experiment will go. Another downside of using a GUI is that there is a risk that reproducibility will suffer if users do not have a mechanism to document their mouse clicks. This documentation is explicit in mothur as all commands and output is recorded in a logfile. Given the heightened recent focus on reproducibility we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript with figures through the Riffomonas project. Finally, we collaborated with programmers through Google Summer of Code to develop commands in mothur to implement the random forest and SVM machine learning algorithms. Similar to the challenges of developing attractive visuals, fitting the algorithms' hyperparameters, testing, and deploying the resulting models requires a significant amount of customization. Furthermore, this is an active area of research where methods are still being developed and improved. Thankfully, there are numerous R and Python packages that do a better job of developing these models. Again, we have put our efforts into developing instructional materials that mothur users can use to fit such models to their data. In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do. In hindsight, our solution to these failure has been to provide tutorials to as a conduit between mothur and their goals.

@@

Beyond the large packages like mothur and QIIME, there has been significant growth in stand alone boutique software tools for sequence curation (e.g. PyroNoise and DADA), chimera checking (e.g. UCHIME, ChimeraSlayer, and Perseus), and clustering (e.g. USEARCH and Swarm). Where possible, we have implemented these algorithms directly into mothur. We have also used this diversity of methods to perform head-to-head comparisons. Most notable is the area of clustering algorithms where there have been a large number of algorithms developed without an obvious method to objectively compare them. Through a series of studies, we applied an objective metric, the Matthew's Correlation Coefficient (MCC), to numerous algorithms for clustering sequences into OTUs. By performing this type of analysis, we were able to objectively compare the algorithms, make recommendations to the field, and develop new algorithms that outperformed the existing algorithms. We have embraced the competition and diversity of all methods being used to analyze amplicon data as an opportunity to identify the strengths and weaknesses of the methods in an attempt to make recommendations to other researchers.

What are we proud of?
* Open source packages
* Platform independent
* Development of open instructional materials
* Data driven development of methods
* Database independent - allows people outside of our field to use mothur
* Reproducibility
* Helped create standards - when something is this popular, you force people to use a certain suite of tools and methods

**The future.** I plan to continue developing mothur for as long as other researchers find it useful. One challenge of such a plan is maintaining the funding to support its development. The development of mothur was initially enabled by a grant to Mitch Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative. We used that seed funding to secure an NSF grant and then a grant from NIH for tool development as part of their Human Microbiome Project. Since that project expired in 201X, we have not had funding to specifically support mothur's development. I have been fortunate to have discretionary funds generated from other projects to help support mothur. Although there is funding for new tools, there appears to be little appetite by funders to support existing tools. Emblematic of this was the NIH program, Big Data To Knowledge (BD2K), which solicited proposals through the program announcement "Extended Development, Hardening and Dissemination of Technologies in Biomedical Computing, Informatics and Big Data Science (PA-14-156)". This opportunity appeared perfect, except that the National Institute of Allergy and Infectious Diseases (NIAID), the primary supporter of microbiome research at NIH, did not participate in the announcement. Tools like mothur have been successful and need mechanisms to continue to mature and support the needs of the research community.

As with anything in science, methods become passe. When we first developed mothur, tools like T-RFLP and DGGE were still commonly used. Today it would be hard to argue that data from those methods meaningfully advance a study relative to what one could get using 16S rRNA gene sequence data. Looking forward, many want to claim that amplicon sequencing is today's DGGE. They claim that instead researchers should move on to shotgun metagenomic sequence data. It is important to note that the two methods answer fundamentally different questions. Metagenomic sequence data tells a researcher about the functional potential and genetic diversity of a community and 16S rRNA gene sequence data describes the taxonomic composition. Both tools provide important information but they cannot easily replace each other. Although metagenomic data does provide highly resolved taxonomic information, the limit of detection is much higher than that of amplicon data. For example, we analyzed 10,000 16S rRNA sequences from each of about 500 subjects. We can think of this as representing about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500 subjects / 5 16S rRNA gene sequences/genome). Assuming a genome is 4 Mbp, this would represent a sequencing depth of 4 Tbp. Although such a sequencing effort is possible, albeit expensive, analyzing such a large dataset with an approach that captures the full genetic diversity of the community would be financially and technically prohibitive. Going forward, there is still a place for 16S rRNA gene sequencing. Although sequencing technologies will continue to evolve to capture longer and more high quality data, there will likely always be a need for characterizing the taxonomic diversity of microbial communities. With this in mind, there will always be a place for tools like mothur.

Of course this does not mean that such tools will remain static. We see three key areas that we will continue to help to develop. First, just as we adapted through the transitions from Sanger to 454 to MiSeq sequencing platforms, we must learn how to adapt data from PacBio, Oxford Nanopore, and other tools to our analysis pipelines. As with these earlier platforms, we must understand the error profiles so that they can be corrected. We have learned that moving forward requires that we maintain or improve sequence quality. No doubt, datasets and read lengths will improve. Second, with these improvements, we will need to continue to improve our algorithms. We have already seen that attempts to use low quality MiSeq and HiSeq data causes computational problems leading to the creation of open and closed reference clustering methods. Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods. More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing quality. Finally, there are ongoing controversies that need further exploration. These include the validity and utility of amplicon sequence variants, the validity and impact of removing low frequency sequences, and methods of identifying and removing 16S rRNA gene sequences from contaminants. In the original mothur paper, we commented that 16S rRNA gene sequencing and analysis is very much like the Red Queen in Alice in Wonderland. Although some disagreed with this analogy, we still feel it fits. The sequencing technology and rapacious appetite of researchers continues to race on. At the same time, bioinformatics tools must adapt to facilitate our research. I am confident that mothur will be up to this exciting challenge.

>>>>

## Abstract
250 words

@@

@@

***Failed experiments.*** If we never failed, we would not be trying hard enough. Over the past decade we have tried a number of experiments to improve the usability and utility of mothur. One of our first experiments was to use mothur to generate standard vector graphic (SVG)-formatted files of heatmaps and venn diagrams depicting the overlap between microbial communities. I quickly realized that I would never put a mothur-generated figure into a manuscript I wrote. Such visuals require far too much customization to be publication-quality. Although QIIME has incorporated visualization tools through the Emperor package, the challenge of users taking default values has downsides as ordinations with black background or publishing 3-D ordinations in a 2-D medium litter the literature. Instead, we have encouraged users to use R packages to visualize mothur-generated results using the minimalR instructional materials. A second experiment we attempted was to create a graphical user interface (GUI) for running mothur. Forcing users to interact with mothur through the command line has been a significant hurdle for many. Unfortunately, the development effort required to create and maintain a GUI is significant and there is limited funding for such efforts. The newest version of QIIME (version 2) has emphasized interaction with the tools through a GUI and it remains to be seen how this experiment will go. Another downside of using a GUI is that there is a risk that reproducibility will suffer if users do not have a mechanism to document their mouse clicks. This documentation is explicit in mothur as all commands and output is recorded in a logfile. Given the heightened recent focus on reproducibility we have extended significant effort in developing instructional materials teaching users how to organize, document, and execute reproducible pipelines that allow a user to go from raw sequence data to a compiled manuscript with figures through the Riffomonas project. Finally, we collaborated with programmers through Google Summer of Code to develop commands in mothur to implement the random forest and SVM machine learning algorithms. Similar to the challenges of developing attractive visuals, fitting the algorithms' hyperparameters, testing, and deploying the resulting models requires a significant amount of customization. Furthermore, this is an active area of research where methods are still being developed and improved. Thankfully, there are numerous R and Python packages that do a better job of developing these models. Again, we have put our efforts into developing instructional materials that mothur users can use to fit such models to their data. In each of our "failed" experiments, the real problems were straying from what mothur does well and failing to grasp what we really wanted the innovation to do.

@@

Beyond the large packages like mothur and QIIME, there has been significant growth in stand alone boutique software tools for sequence curation (e.g. PyroNoise, DADA), chimera checking (e.g. UCHIME, ChimeraSlayer, Perseus), and clustering (e.g. USEARCH, Swarm). Where possible, we have implemented these algorithms directly into mothur. We have also used this diversity of methods to perform head-to-head comparisons. Most notable is the area of clustering algorithms where there have been a large number of algorithms developed without an obvious method to objectively compare them. Through a series of studies, we applied an objective metric, the Matthew's Correlation Coefficient (MCC), to numerous algorithms for clustering sequences into OTUs. By performing this type of analysis, we were able to objectively compare the algorithms, make recommendations to the field, and develop new algorithms that outperformed the existing algorithms. We have embraced the competition and diversity of all methods being used to analyze amplicon data as an opportunity to identify the strengths and weaknesses of the methods in an attempt to make recommendations to other researchers.

***mothur's core principles.*** As mothur has evolved with the needs of the community and matured, several core principles have emerged that direct its development. First, mothur is a free, open source software package. This has been critical in shaping the direction of mothur. We were content for mothur to be an improved combination of DOTUR and SONS and leverage existing tools for other steps. Yet, when we learned that the code for NAST, the algorithm behind the greengenes's aligner,  was not open source or publicly available, we realized that such an important tool needed to be opened to the community to make the tool adaptable to other databases and potential improvements. More recently, the rejection of closed source tools such as USEARCH can be seen by the broader adoption of its competitor, VSEARCH, within the microbial ecology community. Related to insuring that mothur's code is open source, our second core principle is that we maintain transparency for our users. Perhaps a user does not need to interrogate every line of code, but we do feel that they need to understand what is happening. Many programs including QIIME and DADA2 encapsulate large elements of a pipeline in a single command. In contrast, mothur forces the user to specify each step of the pipeline. Although the former approach makes an analysis easier for a beginner, it potentially stifles users that need greater control or understanding of the assumptions for each element of the pipeline. This control over the pipeline has made it easier for researchers to customize databases or adapt the pipeline to analyze non 16S rRNA gene sequence data. Third, as I mentioned above, there has been a plethora of methods proposed for generating amplicon sequence data, and curating, aligning, checking for chimeras, classifying and clustering the data. I am proud of the data-driven approach we have taken to comparing these methods. A description of a new method is of limited value if it is not benchmarked against other methods or control datasets. Through this core principle and mothur's large reach into the community, we have helped to develop standards in the analysis of 16S rRNA gene sequence data. Fourth, a focus on enabling reproducibility has always been central to the functionality of mothur. From the beginning, mothur's logfiles have represented a transcript of the user's command and outputs. When researchers were reluctant to submit sequence data to the Sequence Read Archive (SRA), we worked with the SRA developers to create a mothur command (make.sra) to create the package to submit sequence data through a special mothur portal. A more ambitious project had its seed on April 1, 2013 when we announced a new "function" in mothur: write.paper. The new command required that the user provide a 454 sff file and a journal title or impact factor. With this information, mothur would generate a manuscript. This April Fools joke was poking fun at software that provided an analysis black box but also at the sentiment that data analysis should be so cut and dry. A few years later, we revisited this concept in the scope of reproducibility. Why not explicitly script an analysis from downloading data from the Sequence Read Archive (SRA) through the rendering of a manuscript ready for submission? This idea gave rise to the development of the Riffomonas reproducible research tutorial series that enables researchers to write their own write.paper program. Perhaps the most important core principle is that my research group uses mothur to analyze the data we generate. We "eat our own dogfood". This has proven critical as it again represents transparency to our users and hopefully provides confidence to users that we are not making recommendations that we do not follow ourselves.

***The future.*** Buoyed by these principles, I plan to continue developing mothur for as long as other researchers find it useful. One challenge of such a plan is maintaining the funding to support its development. The development of mothur was initially enabled by a grant to Mitch Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative. We used that seed funding to secure an NSF grant and then a grant from NIH for tool development as part of their Human Microbiome Project. Since that project expired in 201X, we have not had funding to specifically support mothur's development. I have been fortunate to have discretionary funds generated from other projects to help support mothur. Although there is funding for new tools, there appears to be little appetite by funders to support existing tools. Emblematic of this was the NIH program, Big Data To Knowledge (BD2K), which solicited proposals through the program announcement "Extended Development, Hardening and Dissemination of Technologies in Biomedical Computing, Informatics and Big Data Science (PA-14-156)". This opportunity appeared perfect, except that the National Institute of Allergy and Infectious Diseases (NIAID), the primary supporter of microbiome research at NIH, did not participate in the announcement. Tools like mothur have been successful and need mechanisms to continue to mature and support the needs of the research community.

As with anything in science, methods become passe. When we first developed mothur, tools like T-RFLP and DGGE were still commonly used. Today it would be hard to argue that data from those methods meaningfully advance a study relative to what one could get using 16S rRNA gene sequence data. Looking forward, many want to claim that amplicon sequencing is today's DGGE. They claim that researchers should instead move on to shotgun metagenomic sequencing. It is important to note that the two methods answer fundamentally different questions. Metagenomic sequence data tells a researcher about the functional potential and genetic diversity of a community and 16S rRNA gene sequence data describes the taxonomic composition. Both tools provide important information but they cannot easily replace each other. Although metagenomic data does provide highly resolved taxonomic information, the limit of detection is much higher than that of amplicon data. For example, we analyzed 10,000 16S rRNA sequences from each of about 500 subjects. We can think of this as representing about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500 subjects / 5 16S rRNA gene sequences/genome). Assuming a genome is 4 Mbp, this would represent a sequencing depth of 4 Tbp. Although such a sequencing effort is possible, albeit expensive, analyzing such a large dataset with an approach that captures the full genetic diversity of the community would be financially and technically prohibitive. Going forward, there is still a place for 16S rRNA gene sequencing. Although sequencing technologies will continue to evolve to capture longer and more high quality data, there will likely always be a need for characterizing the taxonomic diversity of microbial communities. With this in mind, there will always be a place for tools like mothur.

Of course this does not mean that such tools will remain static. We see three key areas that we will continue to help to develop. First, just as we adapted through the transitions from Sanger to 454 to MiSeq sequencing platforms, we must learn whether data from PacBio and Oxford Nanopore are acceptable alternatives and then our analysis pipelines to facilitate their analysis. As with these earlier platforms, we must understand the error profiles so that they can be corrected. We have learned that moving forward requires that we maintain or improve sequence quality. No doubt, datasets and read lengths will improve. Second, with these improvements, we will need to continue to improve our algorithms. We have already seen that attempts to use low quality MiSeq and HiSeq data causes computational problems leading to the creation of open and closed reference clustering methods. Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods. More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing quality. Finally, there are ongoing controversies that need further exploration. These include the validity and utility of amplicon sequence variants, the validity and impact of removing low frequency sequences, and methods of identifying and removing 16S rRNA gene sequences from contaminants. In the paper announcing mothur we commented that the relationship between 16S rRNA gene sequencing and analysis is very much like the Red Queen in Alice in Wonderland. Although some disagreed with this analogy, we still feel it fits. The sequencing technology and rapacious appetite of researchers continues to race on. At the same time, bioinformatics tools must adapt to facilitate our research. I am confident that mothur will be up to this exciting challenge.

***Conclusion.*** Beyond its utility for analyzing amplicon sequence data, I believe that mothur's history provides lessons that are helpful for other projects with a long historical arc. First, mothur is a product of its time. It has always sought to solve a current need to the best of our ability with the tools we had at the time. There are certainly caveats to any analysis of 16S rRNA gene sequence data, but if we had waited until those caveats were resolved, the field never would have progressed. Similarly, we made design choices that we probably would not have made had we started the project today. Second, as we have developed mothur, we have attempted to do so in a data driven approach where we compare multiple competing methods to each other. It has not merely been enough to propose a new method, we must show that it meaningfully advances the field. Third, through our failures and successes we have learned to focus on what mothur is good at and create alternative products to help users. For example, mothur will not have a graphical interface. Instead, we will provide instructional materials to teach users how to use the command line interface and other programs like R to generate a reproducible workflow. Finally, mothur was born out of a need for automating the analysis of large 16S rRNA gene sequence datasets. I realized I had a set of skills to fill that need. It has been refreshing to see the computational skills of the microbial ecology field grow over the past decade. Looking ahead, we must all take stock of the challenges we face in microbial ecology and how our unique skills and interests can address these challenges to turn them into opportunities.
